### 2020-12-18 Thesis-Chapter2-Background.tex
Update forms, small editorial tweaks

The problem of strong encryption and \acl{EA} exists in a technical, historical, and regulatory context. This chapter
introduces cryptography, summarizes the history of encryption and the ``crypto wars,'' lists prominent technical \ac{EA}
proposals, and lastly overviews relevant laws and regulations.

@@

There are two major cryptographic protocol families, symmetric and asymmetric. Symmetric protocols have been around for
thousands of years, but asymmetric protocols were only invented in recent decades. In \ac{sym-crypto}, encryption and
decryption are performed with the same key, and both parties must have this key in order to communicate securely. In
\ac{asym-crypto}, also known as \ac{public-key-crypto}, encryption and decryption are performed with two paired keys,
called the public key and the private key. The public key is not secret, but due to its relationship with the private
key, it can be used to establish identity and encrypted communication. Using this technique, two parties can communicate
privately without requiring previously agreed-upon secret information.

There are also two major cryptographic applications, securing \ac{DIM} and \ac{DAR}. Each application presents different
challenges that require different solutions. \Ac{DIM} typically uses long-lived \ac{asym-crypto} keys to perform
authentication and establish short-lived \ac{sym-crypto} session keys. The session keys perform the actual encryption of
the \acl{DIM}. \Ac{forward-secrecy} and \ac{replay-protection} are two important properties of \ac{DIM} encryption
protocols. \Ac{forward-secrecy} ensures that a leaked private key or session key does not compromise any other private
key or session key; \ac{replay-protection} ensures that messages cannot be replayed by an attacker without detection
\cite{bellovin_thinking_2016}. \Ac{E2EE} for instant messaging services is an example of encryption for \ac{DIM}.

\Ac{DAR} by nature must use long-lived keys for encryption. Instead of being negotiated and randomly generated at
encryption-time, as \ac{DIM} session keys are, these keys are either derived from a user-entered password or are stored
somewhere in computer memory. Secure storage often takes place with the assistance of dedicated hardware.
\Ac{disk-encryption} for laptops and mobile devices is an example of encryption for \ac{DAR}.

@@

Cryptography is the technical foundation for many forms of computer and network security. Security is defined in terms
of several properties: authentication, integrity, non-repudiation, confidentiality, availability, and authorization
\cite{shostack_threat_2014}. These system-level properties emerge from both the composition of the system's components
(architecture) and the security of the components themselves. Likewise, the security of each component emerges from the
architecture of its subcomponents and the security of the subcomponents themselves. At the end of this chain of
analysis, the security of primitive components commonly relies on cryptography.

@@

Violations of cryptographic integrity in primitive components, introduced by design or by accident, could have
catastrophic effects. Compromised digital certificates, hashes, or encryption would enable spoofing, tampering, and
information disclosure done both for their own sake and as tools in general attacks. The importance of neutralizing
these tools is clear when such attacks include interference with elections \cite{mueller_2018}, multi-billion dollar
disruptions of business \cite{greenberg_2018_notpetya}, and hospital network ransoms that disrupt care of patients
\cite{goodin_ransomware_2019}.

@@

Encryption's dual role as an enabler of privacy and cornerstone of security is at the heart of the EA debate.
Cryptography has an absolute, mathematical power---a power that is objectionable in the realm of privacy, but necessary
in the realm of security.

@@

The section provides a brief overview of encryption from ancient to modern times.

@@

Computerized encryption is a new technology, but the field of cryptography is old, since demand for privacy is as old as
communication itself. The most well-known ancient example of rudimentary encryption is the Caesar cipher, named after
the character substitution technique Julius Caesar used to protect private correspondence
\cite{luciano_cryptology_1987}. Significant developments include the first formal cryptographic study by Arab scholars
in the eighth century, advancements made out of necessity in the twentieth century's world wars, and the application of
computers to cryptographic problems \cite{kahn_codebreakers_1996}. Claude Shannon formalized the modern ``mathematical
analysis of cryptography'' in 1949 \cite{shannon_communication_1949} and Whitfield Diffie and Martin Hellman published
research on \ac{public-key-crypto} in 1976 \cite{diffie_new_1976}. The discovery of \ac{public-key-crypto} was an
important advance. Combined with computer networking and personal computing advances in the 1980s, it put the power of
strong encryption not just in the hands of governments and militaries, but of ordinary people.

@@

In 1976, Diffie and Hellmann published their seminal research in public key cryptography. That same year, the U.S.
Congress passed the \ac{AECA} and declared that strong cryptography is subject to export controls
\cite{kehl_right_2015}. \Ac{the-cw1} began in 1991, when the U.S. Senate introduced, but did not pass, a bill mandating
access to plain text contents when authorized by law. In response, Philip Zimmermann released \ac{PGP}, e-mail
encryption software, in order that strong cryptography be ``made available to the American public before it became
illegal to use'' \cite{zimmermann_1996}. In 1993, the Clinton administration introduced the \ac{clipperchip}
\cite{press_1993} with the goal of ``providing the public with strong cryptographic tools without sacrificing the
ability of law enforcement and intelligence agencies to access unencrypted versions of those communications''
\cite{thompson_2015}. Citing the foundational role of cryptography in security---and the potential of human rights
abuses for compromised privacy---industrial and technical leaders reacted negatively to the initiative
\cite{kehl_right_2015} \cite{zimmermann_1996}. When a prominent security researcher discovered flaws that allowed users
to subvert the \ac{clipperchip} mechanisms \cite{blaze_protocol_1994}, the proposal died.

Despite the failure of the \ac{clipperchip}, the debate over export controls and access to strong encryption continued,
primarily focused on various key escrow proposals \cite{thompson_2015}. In 1996, two pro-encryption bills were
introduced in the U.S. Congress, one in the Senate and one in the House. The Senate saw S.1726, the \ac{PROCODEAct} of
1996, which sought to abolish the export controls on encryption software \cite{burns_s1726_1996}. The House saw HR.3011,
the \ac{SAFEAct} of 1996, which similarly sought to remove export controls, but further sought to explicitly allow
arbitrarily strong encryption for all legal activity and preclude mandatory \ac{EA} schemes such as the \ac{clipperchip}
\cite{goodlatte_hr3011_1996}. Zimmermann, the author of PGP, who had by this time endured an investigation by the
Customs Service for publishing his work, testified before the Senate in favor of the \ac{PROCODEAct}
\cite{zimmermann_1996}. In November of 1996, the Clinton administration released an executive order effectively removing
export controls on encryption products along the lines of the \ac{PROCODEAct}, which never came to pass
\cite{clinton_1996}.

From 1996 to 1999, the \ac{SAFEAct} was proposed several times, was discussed at hearings, and gained support
\cite{kehl_right_2015}. By this time, there was ``an overwhelming amount of evidence against moving ahead with any key
escrow schemes'' \cite{thompson_2015}, and in 1999 the Clinton administration abruptly changed course, adopting almost
all \ac{SAFEAct} proposals \cite{kehl_right_2015}. This development marked the end of \ac{the-cw1}.

@@

Two developments paved the way for \ac{the-cw2}. The first was the result of inaction: when the Clinton administration
changed its encryption policy, the U.S. House dropped the \ac{SAFEAct} \cite{goodlatte_hr3011_1996}. The \ac{PROCODEAct}
and other encryption-related bills went unpassed as well, as elaborated in \mysec{sec-us-reg-fails}. This legislative
failure meant that rather than being written in the pen of law, encryption policy was written in the pencil of executive
order.

@@

\paragraph*{``Security vs. Security''} The relationship between security and privacy is sometimes viewed as
antagonistic, an assumption implicit in the ``nothing-to-hide'' argument. However, discussions in recent years have
shown that the debate is more about ``Security vs. Security'' than ``Security vs. Privacy''
\cite{stalla_bourdillon_privacy_2014}. In this context, there are two types of security. First, public security (often
referred to as ``safety'' in this paper), which is the pursuit of law enforcement. Second, cybersecurity, which the
pursuit of the technical community. Both sides of the debate have by now acknowledged that privacy and public security
are not always in conflict. In fact \ac{EA}'s effects on public security vs. cybersecurity---which in return affects
public security---may be more important \cite{schneier_2019}.

\paragraph*{Apple vs. FBI} The 2015 San Bernardino terror attack resulted in the first major battle of \ac{the-cw2}. The
suspect was killed, leaving behind a locked iPhone. The \ac{FBI} issued an order under the \acrlong{AWA} to compel Apple
to unlock the device, which was among the first generation of Apple's fully encrypted iPhones. Apple objected on grounds
that it was ``unreasonably burdensome'' and would undercut the integrity of all iOS devices \cite{schulze_clipper_2017}.
The case occupied U.S. District courts, the media, and the attention of the political world from February 16, 2016, when
the warrant was issued, to March 28, when the \ac{FBI} announced they had gained access to the phone through alternate
means \cite{novet_2016}. Although the phone proved not to contain any important data \cite{schulze_clipper_2017}, the
high profile case featured two characteristic elements of the new crypto wars---terrorism and device encryption. Shortly
after the case ended, a pair of Republican and Democratic senators jointly released a discussion draft of a bill that
would have forced Apple to comply \cite{burr_2016}. The draft did not make it to the floor, and the unresolved nature of
the court case left the legal debate unresolved.

@@

\paragraph*{Volatile Politics} Finally, \ac{the-cw2} is taking place amongst a political landscape marked by
uncertainty, extremism, misinformation, and a pandemic that will have unknown long-term impacts. Since government is
part of the broader data security threat model, its behavior is important. Unfortunately, it is difficult for the
technical community to engage productively with policymakers due to present political dynamics and the lack of reliable
data showing what law enforcement agencies need \cite{granick_2018}.

That brings us to where the debate stands today. A December 2019 U.S. Senate Judiciary Committee featured the top two
U.S. parties both expressing anger over whether data should be ``beyond the reach of the law.'' Senators threatened tech
executives with statements such as ``get your act together, or we will gladly get your act together for you''
\cite{geller_2019}. In March 2020, the Republican atop the Senate Judiciary Committee introduced S.3398, the
\ac{EARNITAct} of 2020, which seeks to establish a 16-chair committee with the power to revoke online platforms'
liability protection for user-submitted content if the committee determines the platform is not doing enough to fight
\ac{CSAM} \cite{graham_s3398_2020}. As illustrated by the previously mentioned \textit{New York Times} feature
\cite{keller_internet_2019} that likely triggered the December 2019 hearing, the problem is enormous. However, critics
largely saw the \ac{EARNITAct} as an attempt to indirectly yet effectively outlaw \ac{E2EE} \cite{newman_2020}
\cite{pfefferkorn_2020}. In response, legislators narrowed the focus of the \ac{EARNITAct}, but only after unveiling the
\ac{LAEDAct}

@@

Any scheme that relies on the fidelity and security of information held by an entity besides the data owner (in the case
of \ac{DAR}) or sender or receiver (in the case of \ac{DIM}) is a ``trusted-party'' key escrow approach. Though there is
a wide variety of implementation options within this category \cite{denning_taxonomy_1996}---secret or open protocol,
hardware-enabled or software-only, single key or split key, government or third-party escrow---the foundation of the
effectiveness of these approaches lies in the security of the trusted party. The \ac{clipperchip} is one example, being
a secret protocol hardware-enabled split key government escrow system \cite{denning_taxonomy_1996}.

@@

The third type of \ac{key-escrow} is device-based. In device-based \ac{key-escrow}, special key information is held not
by trusted parties or distributed systems, but on the device itself in specialized hardware. Advances in secure hardware
``enclaves'' since \ac{the-cw1} have enabled secure local key storage. Approaches relying on hardware primarily target
\ac{DAR} and may include time elements and split keys with trusted parties as well \cite{savage_lawful_2018}
\cite{ozzie_2018}.

@@

\Ac{lawful-hacking} is one way to access \ac{plaintext} without \ac{EA}. Under \ac{lawful-hacking}, instead of having
access to plaintext data via an alternative means of decryption, law enforcement is allowed to perform otherwise-illegal
hacking activity in order to find the key, compromise a device, or intercept plaintext versions of the data
\cite{kerr_encryption_2017}. While some see this as a viable middle ground \cite{nguyen_lawful_2017}, encouraging law
enforcement to exploit vulnerabilities creates mis-aligned incentives. If law enforcement opposes strong security in
general, it could foster the exploit market while costing ordinary users money and security \cite{soesanto_2018}.

@@

Simpler methods than computer hacking or legal coercion can be used: encrypted data may exist in plaintext elsewhere, or
authorities may seize equipment while it is unlocked \cite{kerr_encryption_2017}. This approach relies on law
enforcement's skills in investigation and sting operations. This approach is used successfully today. In 2017, an
international law enforcement operation used this technique to take down a large illicit online drug market
\cite{greenberg_2018}. This approach also does not address \ac{DIM} and requires well-executed operations.

@@

\acs{ECPA} includes the Stored Communications Act and Pen Register Act, amending the switchboard-era Wiretap Act.
Provides explicit protections for private electronic communications data, and updated definitions to contemporary
technology. It grants law enforcement explicit access to certain data as well; some terminology has aged poorly with
technological evolution, making the access increasingly permissive \cite{shamsi_2011}.

@@

Written in the aftermath of September 11, the \ac{USAPATRIOTAct} broadly expanded U.S. government surveillance powers in
25 sections of ``Title II: Enhanced Surveillance Procedures.'' The act notably expands \acs{FISA} order scope (Section
206), duration (207), and use on U.S. citizens (218). Section 215 enables court orders requiring business records, but
the Snowden leaks revealed that the government had been secretly using a non-standard definition of ``business records''
in order ``to justify requests for domestic telephone metadata delivered in bulk, not individualized requests''
\cite{landau_making_2013}.

@@

The \ac{LAEDAct} is a direct attack on encryption and the direct successor of the Compliance with Court Orders Act. It
was introduced by the chair of the Senate Judiciary Committee, also sponsor of the \ac{EARNITAct} and one of the
outspoken encryption critics during the 2019 Senate hearing on encryption and \ac{EA}. The \ac{LAEDAct} specifically
addresses \ac{DAR} and \ac{DIM} by forcing device, operating system, and application developers to implement the ability
to decrypt all data stored or passing through the device or software system ``concurrently with their transmission''
``unless the independent actions of an unaffiliated entity make it technically impossible to do so.'' The act neither
mandates nor recommends any technical approach to providing this capability.

@@

Judicial influence on encryption policy is primarily mediated through the Fourth Amendment, particularly regarding the
interpretation of a person's ``effects,'' what is ``reasonable,'' and what defines a ``search.'' The Supreme Court's
interpretation and implementation of the Amendment has changed over time. Legal scholars Craig Curtis and Michael Gizzi
conducted an in-depth analysis of the Court's jurisprudence from the 1960s Warren Court, through the influential 1980s
Rehnquist Court, to the mid-2010s Roberts Court \cite{gizzi_fourth_2016}. Their analysis revealed a deep proclivity
towards crime control that is only recently thawing. This gradual return to decisions favoring civil liberties, caused
by evolution of technology, of the court, and of the justices themselves, has been favorable for data privacy. Recent
cases have increasingly sided with defendants, caused justices to strain to show that they understand technology, and
increased the categories of protected digital data \cite{gizzi_fourth_2016}. 2018's potentially landmark \ii{Carpenter
v. United States} may be the most impactful yet. Though the case only resulted in a narrow decision against unwarranted
cell phone location data tracking, the Court's decision undermines the pre-digital third-party doctrine, which holds
that Fourth Amendment protection is forfeit when the data is willingly given to a third-party \cite{franklin_2018}.

As this relates to \ac{EA}, Fourth Amendment interpretation determines the threshold at which authorities require a
warrant, not what demands they can or cannot make once they have one. As mentioned above, the FBI argued in the San
Bernardino case that the \ac{AWA} grants it authority to demand access, though others argue that \acs{CALEA} explicitly
precludes this kind of demand \cite{gidari_2016}. Since the FBI dropped the case, the matter is not formally settled,
though the consensus is that legislation such as the \ac{LAEDAct} is required before courts would uphold that command.

@@

That judicial influence on privacy and technology is changing is to be expected. To explain the fluctuating state of
legal interpretation as it relates to \ac{EA}, Matyas et al. make the helpful distinction between legal rules and
principles \cite{matyas_incommensurability_2018}:

@@

In 1946 The U.S. formalized an intelligence-sharing operation with the U.K. in the BRUSA (now UKUSA) Agreement. Over the
following 10 years, Australia, Canada, and New Zealand were made full partners to this agreement, which has become known
as ``Five Eyes'' \cite{nsa_2020}. Due to the close relationship between the intelligence and law enforcement agencies of
these countries, the member's encryption policies are relevant to one another.

>>>>

The problem of strong encryption and \acl{EA} exists in a technical, historical, and regulatory context. This chapter
introduces cryptography, summarizes the history of encryption and the ``crypto wars,'' lists prominent technical \ac{EA}
proposals, and overviews relevant laws and regulations.

@@

There are two major cryptographic protocol families, symmetric and asymmetric. Symmetric protocols have been around for
thousands of years, but asymmetric protocols were only invented in recent decades. In \ac{sym-crypto}, encryption and
decryption are performed with the same key, and both parties must have this key in order to communicate securely. In
\ac{asym-crypto}, also known as \ac{public-key-crypto}, encryption and decryption are performed with two paired keys,
called the public key and the private key. The public key is not secret, but due to its relationship with the private
key, it can be used to establish identity and initiate encrypted communications. Using this technique, two parties can
communicate privately without requiring previously agreed-upon secret information.

There are also two major cryptographic applications, securing \ac{DIM} and \ac{DAR}. Each application presents different
challenges that require different solutions. \Ac{DIM} typically uses long-lived \ac{asym-crypto} keys to perform
authentication and to establish short-lived \ac{sym-crypto} session keys. The session keys perform the actual encryption
of the \acl{DIM}. \Ac{forward-secrecy} and \ac{replay-protection} are two important properties of \ac{DIM} encryption
protocols. \Ac{forward-secrecy} ensures that a leaked private key or session key does not compromise any other private
key or session key; \ac{replay-protection} ensures that messages cannot be replayed by an attacker without detection
\cite{bellovin_thinking_2016}. \Ac{E2EE} for instant messaging services is an example of encryption for \ac{DIM}.

\Ac{DAR} by nature must use long-lived keys for encryption. Instead of being negotiated and randomly generated at
encryption-time, as \ac{DIM} session keys are, these keys are either derived from user-entered passwords or are stored
somewhere in computer memory. Secure storage often takes place with the assistance of dedicated hardware.
\Ac{disk-encryption} for laptops and mobile devices is an example of encryption for \ac{DAR}.

@@

Cryptography is the technical foundation for many forms of computer and network security. Security is defined in terms
of several properties: authentication, integrity, non-repudiation, confidentiality, availability, and authorization
\cite{shostack_threat_2014}. These system-level properties emerge from both the composition of the system's components
(architecture) and the security of the components themselves. Likewise, the security of each component emerges from the
architecture of its subcomponents and the security of the subcomponents themselves. At the bottom of this chain of
analysis, the security of primitive components commonly relies on cryptography.

@@

Violations of cryptographic integrity in primitive components, introduced by design or by accident, could have
catastrophic effects. Compromised digital certificates, hashes, or encryption would enable spoofing, tampering, and
information disclosure attacks. The attacks could be done for their own sake or as steps in larger attack chains. The
importance of neutralizing these attacks is clear when they result in interference with elections \cite{mueller_2018},
multi-billion dollar disruptions of business \cite{greenberg_2018_notpetya}, and hospital network ransoms that disrupt
care of patients \cite{goodin_ransomware_2019}.

@@

Cryptography has an absolute, mathematical power---a power that is necessary in the realm of security, but objectionable
in the realm of privacy. Encryption's dual role as an enabler of privacy and cornerstone of security is at the heart of
the EA debate.

@@

The section provides a brief history of encryption from ancient to modern times.

@@

Computerized encryption is a new technology, but the field of cryptography is old, since demand for privacy is as old as
communication itself. The most well-known ancient example of rudimentary encryption is the Caesar cipher, named after
the character substitution technique Julius Caesar used to protect private correspondence
\cite{luciano_cryptology_1987}. Significant developments include the first formal cryptographic study by Arab scholars
in the eighth century, advancements made out of necessity during the twentieth century's world wars, and the application
of computers to cryptographic problems \cite{kahn_codebreakers_1996}. Claude Shannon formalized the modern
``mathematical analysis of cryptography'' in 1949 \cite{shannon_communication_1949} and Whitfield Diffie and Martin
Hellman published research on \ac{public-key-crypto} in 1976 \cite{diffie_new_1976}. The discovery of
\ac{public-key-crypto} was an important advance. Combined with computer networking and personal computing advances in
the 1980s, it put the power of strong encryption not just in the hands of governments and militaries, but of ordinary
people.

@@

In 1976, Diffie and Hellmann published their seminal research in public key cryptography. That same year, the U.S.
Congress passed the \ac{AECA} and declared that strong cryptography is subject to export controls
\cite{kehl_right_2015}. \Ac{the-cw1} began in 1991, when the U.S. Senate introduced, but did not pass, a bill mandating
access to plain text contents when authorized by law. In response, Philip Zimmermann released \ac{PGP}, e-mail
encryption software, in order that strong cryptography be ``made available to the American public before it became
illegal to use'' \cite{zimmermann_1996}. In 1993, the Clinton administration introduced the \ac{clipperchip}
\cite{press_1993} with the goal of ``providing the public with strong cryptographic tools without sacrificing the
ability of law enforcement and intelligence agencies to access unencrypted versions of those communications''
\cite{thompson_2015}. Citing the foundational role of cryptography in security---and the potential of human rights
abuses resulting from compromised privacy---industrial and technical leaders reacted negatively to the initiative
\cite{kehl_right_2015} \cite{zimmermann_1996}. When a prominent security researcher discovered flaws that allowed users
to subvert the \ac{clipperchip} mechanisms \cite{blaze_protocol_1994}, the proposal died.

Despite the failure of the \ac{clipperchip}, the debate over export controls and access to strong encryption continued,
primarily focused on various key escrow proposals \cite{thompson_2015}. In 1996, two pro-encryption bills were
introduced in the U.S. Congress, one in the Senate and one in the House. The Senate saw S.1726, the \ac{PROCODEAct} of
1996, which sought to abolish the export controls on encryption software \cite{burns_s1726_1996}. The House saw HR.3011,
the \ac{SAFEAct} of 1996, which similarly sought to remove export controls, but further sought to explicitly allow
arbitrarily strong encryption for all legal activity and to preclude mandatory \ac{EA} schemes such as the
\ac{clipperchip} \cite{goodlatte_hr3011_1996}. Zimmermann, the author of PGP, who had by this time endured an
investigation by the Customs Service for publishing his work, testified before the Senate in favor of the
\ac{PROCODEAct} \cite{zimmermann_1996}. In November of 1996, the Clinton administration released an executive order
effectively removing export controls on encryption products along the lines of the \ac{PROCODEAct}, which never came to
pass \cite{clinton_1996}.

From 1996 to 1999, the \ac{SAFEAct} was proposed several times, was discussed at hearings, and gained support
\cite{kehl_right_2015}. By this time, there was ``an overwhelming amount of evidence against moving ahead with any key
escrow schemes'' \cite{thompson_2015}, and in 1999 the Clinton administration changed course again, adopting almost all
\ac{SAFEAct} proposals \cite{kehl_right_2015}. This development marked the end of \ac{the-cw1}.

@@

Two developments paved the way for \ac{the-cw2}. The first was the result of inaction: when the Clinton administration
changed its encryption policy, the U.S. House dropped the \ac{SAFEAct} \cite{goodlatte_hr3011_1996}. The \ac{PROCODEAct}
and other encryption-related bills went unpassed as well, as elaborated in \mysec{sec-us-reg-fails}. This legislative
failure meant that encryption policy was written not in the pen of law, but in the pencil of executive order.

@@

\paragraph*{``Security vs. Security''} The relationship between security and privacy is sometimes viewed as
antagonistic, an assumption implicit in the ``nothing-to-hide'' argument. However, discussions in recent years have
shown that the debate is more about ``Security vs. Security'' than ``Security vs. Privacy''
\cite{stalla_bourdillon_privacy_2014}. In this context, there are two types of security. First, public security (often
referred to as ``safety'' in this paper), which is the pursuit of law enforcement. Second, cybersecurity, which the
pursuit of the technical community. Both sides of the debate have by now acknowledged that privacy and public security
are not always in conflict. In fact, it is the relationship between public security and cybersecurity---which in return
affects public security---that be more important \cite{schneier_2019}.

\paragraph*{Apple vs. FBI} The 2015 San Bernardino terror attack resulted in the first major battle of \ac{the-cw2}. The
attacker was killed, leaving behind a locked iPhone. The \ac{FBI} issued an order under the \acrlong{AWA} to compel
Apple to unlock the device, which was among the first generation of Apple's fully encrypted iPhones. Apple objected on
grounds that it was ``unreasonably burdensome'' and would undercut the integrity of all iOS devices
\cite{schulze_clipper_2017}. The case occupied U.S. District courts, the media, and the attention of the political world
from February 16, 2016, when the warrant was issued, to March 28, when the \ac{FBI} announced they had gained access to
the phone through alternate means \cite{novet_2016}. Although the phone proved not to contain any important data
\cite{schulze_clipper_2017}, the high profile case featured two characteristic elements of the new crypto
wars---terrorism and device encryption. Shortly after the case ended, a pair of Republican and Democratic senators
jointly released a discussion draft of a bill that would have forced Apple to comply \cite{burr_2016}. The draft did not
make it to the Senate floor, and the unresolved nature of the court case left the legal debate unresolved.

@@

\paragraph*{Volatile Politics} Finally, \ac{the-cw2} is taking place amidst a political landscape marked by uncertainty,
extremism, misinformation, and a pandemic that will have unknown long-term impacts. Since government is part of the
broader data security threat model, its behavior is important. Unfortunately, it is difficult for the technical
community to engage productively with policymakers due to present political dynamics and the lack of reliable data
demonstrating the extent of law enforcement's difficulties \cite{granick_2018}.

That brings us to where the debate stands today. A December 2019 U.S. Senate Judiciary Committee featured the top two
U.S. parties both expressing anger over whether data should be ``beyond the reach of the law.'' Senators threatened tech
executives with statements such as ``get your act together, or we will gladly get your act together for you''
\cite{geller_2019}. In March 2020, the Republican atop the Senate Judiciary Committee introduced S.3398, the
\ac{EARNITAct} of 2020, which seeks to establish a 16-chair committee with the power to revoke online platforms'
liability protection for user-submitted content if the committee determines the platform is not doing enough to fight
\ac{CSAM} \cite{graham_s3398_2020}. As illustrated by the previously mentioned \textit{New York Times} feature
\cite{keller_internet_2019} that likely triggered the December 2019 hearing, the problem is enormous. However, critics
largely saw the \ac{EARNITAct} as an attempt to indirectly yet effectively outlaw \ac{E2EE} \cite{newman_2020}
\cite{pfefferkorn_2020}. In response, legislators narrowed the focus of the \ac{EARNITAct}, but only after unveiling the
\ac{LAEDAct}---a direct attack on strong encryption calling for broad \acl{EA} capabilities.

@@

Any scheme that relies on the fidelity and security of information held by an entity besides the data owner (in the case
of \ac{DAR}) or sender or receiver (in the case of \ac{DIM}) is a ``trusted-party'' key escrow approach. Though there is
a wide variety of implementation options within this category \cite{denning_taxonomy_1996}---secret or open protocol,
hardware-enabled or software-only, single- key or split-key, government or third-party escrow---the foundation of the
effectiveness of these approaches lies in the security of the trusted party. The \ac{clipperchip} is one example, being
a secret-protocol hardware-enabled split-key government escrow system \cite{denning_taxonomy_1996}.

@@

The third type of \ac{key-escrow} is device-based. In device \ac{key-escrow}, key information is held not by trusted
parties or distributed systems, but on the device itself in specialized hardware. Advances in secure hardware
``enclaves'' since \ac{the-cw1} have enabled secure local key storage. Approaches relying on hardware primarily target
\ac{DAR} and may include time elements and split keys with trusted parties as well \cite{savage_lawful_2018}
\cite{ozzie_2018}.

@@

\Ac{lawful-hacking} is one way to access \ac{plaintext} without \ac{EA}. Under \ac{lawful-hacking}, instead of having
access to plaintext data via an alternative means of decryption, law enforcement is allowed to perform otherwise-illegal
hacking activity in order to find the key, compromise a device, or intercept plaintext versions of the data
\cite{kerr_encryption_2017}. While some see this as a viable middle ground \cite{nguyen_lawful_2017}, encouraging law
enforcement to exploit vulnerabilities can create misaligned incentives. If law enforcement opposes strong security in
general, it could foster the exploit market while costing ordinary users money and security \cite{soesanto_2018}.

@@

Simpler methods than computer hacking or legal coercion can be used: encrypted data may exist in plaintext copies
elsewhere, or authorities may seize equipment while it is unlocked \cite{kerr_encryption_2017}. This approach relies on
law enforcement's skills in investigation and sting operations. This approach is used successfully today. In 2017, an
international law enforcement operation used this technique to take down a large illicit online drug market
\cite{greenberg_2018}. This approach also does not address \ac{DIM} and requires well-executed operations.

@@

\acs{ECPA} includes the Stored Communications Act and Pen Register Act, and amends the switchboard-era Wiretap Act.
\acs{ECPA} provides explicit protections for private electronic communications data, and updated definitions to
contemporary technology. It grants law enforcement explicit access to certain data as well; some terminology has aged
poorly with technological evolution, making the access increasingly permissive \cite{shamsi_2011}.

@@

Written in the aftermath of September 11, the \ac{USAPATRIOTAct} broadly expanded U.S. government surveillance powers in
the 25 sections of ``Title II: Enhanced Surveillance Procedures.'' The act notably expanded \acs{FISA} order scope
(Section 206), duration (207), and use on U.S. citizens (218). Section 215 enables court orders requiring business
records, but the Snowden leaks revealed that the government had been secretly using a non-standard definition of
``business records'' in order ``to justify requests for domestic telephone metadata delivered in bulk, not
individualized requests'' \cite{landau_making_2013}.

@@

The \ac{LAEDAct} is a direct attack on encryption and the spiritual successor of the Compliance with Court Orders Act.
It was introduced by the chair of the Senate Judiciary Committee, also sponsor of the \ac{EARNITAct} and one of the
outspoken encryption critics during the 2019 Senate hearing on encryption and \ac{EA}. The \ac{LAEDAct} specifically
addresses \ac{DAR} and \ac{DIM} by forcing device, operating system, and application developers to implement the ability
to decrypt all data stored or passing through the device or software system ``concurrently with their transmission''
``unless the independent actions of an unaffiliated entity make it technically impossible to do so.'' The act neither
mandates nor recommends any technical approach to providing this capability.

@@

Judicial influence on encryption policy is primarily mediated through the Fourth Amendment, particularly regarding the
interpretation of a person's ``effects,'' what is ``reasonable,'' and what defines a ``search.'' The Supreme Court's
interpretation and implementation of the Amendment has changed over time. Legal scholars Craig Curtis and Michael Gizzi
conducted an in-depth analysis of the Court's jurisprudence from the 1960s Warren Court, through the influential 1980s
Rehnquist Court, to the mid-2010s Roberts Court \cite{gizzi_fourth_2016}. Their analysis revealed a deep proclivity
towards crime control that is only recently thawing. This gradual return to decisions favoring civil liberties, caused
by evolution of technology, of the court, and of the justices themselves, has been favorable for data privacy. Recent
cases have increasingly sided with defendants, caused justices to strain to show that they understand technology, and
increased the categories of protected digital data \cite{gizzi_fourth_2016}. 2018's potentially landmark \ii{Carpenter
v. United States} may be the most impactful yet. Though the case only resulted in a narrow decision against unwarranted
cell phone location data tracking, the Court's decision undermines the pre-digital era third-party doctrine, which holds
that Fourth Amendment protection is forfeit when the data is willingly given to a third-party \cite{franklin_2018}.

As this relates to \ac{EA}, Fourth Amendment interpretation determines the threshold at which authorities require a
warrant, but does not specify what demands they can or cannot make once they have one. As mentioned above, the FBI
argued in the San Bernardino case that the \ac{AWA} grants it authority to demand access to \ac{plaintext}, though
others argue that \acs{CALEA} explicitly precludes this kind of demand \cite{gidari_2016}. Since the FBI dropped the
case, the matter is not formally settled, though the consensus is that legislation such as the \ac{LAEDAct} is required
before courts would uphold such a command.

@@

It is to be expected that judicial influence on privacy and technology is evolving. To explain the fluctuating state of
legal interpretation as it relates to \ac{EA}, Vashek Matyas et al. make the helpful distinction between legal rules and
principles \cite{matyas_incommensurability_2018}:

@@

In 1946 The U.S. formalized an intelligence-sharing operation with the U.K. in the BRUSA (now UKUSA) Agreement. Over the
following 10 years, Australia, Canada, and New Zealand were made full partners to this agreement, which has become known
as ``Five Eyes'' \cite{nsa_2020}. Due to the close relationship between the intelligence and law enforcement agencies of
these countries, the member's encryption policies are relevant to each another.\\

### 2020-12-18 Thesis-Chapter6-ThreatModel.tex
Update forms, small editorial tweaks

\modelitem{Foreign intelligence}{Foreign intelligence agencies and \acp{APT} are constantly looking for ways to escalate
their privilege and tamper with or steal information. They have world-class expertise and computing power, which they
can apply to both individual devices and backend systems. (Here I am referring to \ii{foreign} foreign intelligence
agencies, as opposed to \ii{domestic} foreign intelligence agencies such as the \acs{CIA} or \acs{NSA} in the U.S.,
which are assumed to operate within---though push the boundaries of---the legal framework.)}

@@

Elevation of privilege are listed in \mytab{table-privilege}. Spoofing to usurp a entity's privilege or using any attack
to unlock a device without authorization can be seen as escalation of privilege.

>>>>

\modelitem{Foreign intelligence}{Foreign intelligence agencies and \acp{APT} are constantly looking for ways to elevate
their privilege and tamper with or steal information. They have world-class expertise and computing power, which they
can apply to both individual devices and backend systems. (Here I am referring to \ii{foreign} foreign intelligence
agencies, as opposed to \ii{domestic} foreign intelligence agencies such as the \acs{CIA} or \acs{NSA} in the U.S.,
which are assumed to operate within---though push the boundaries of---the legal framework.)}

@@

Elevation of privilege are listed in \mytab{table-privilege}. Spoofing to usurp a entity's privilege or using any attack
to unlock a device without authorization can be seen as elevation of privilege.
